---
title: "crops_model"
author: "Vanessa Rathbone"
date: "5/13/2021"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read in the data

- The read.table command loads the CSV file provided by NCDC into a table structure in R, here called 'clim'; the as.Date command then transforms the DATE field in that table into an R-formatted date array that R knows how to do things with.

```{r readdata}
library(lubridate)
library(ggplot2)
library(tidyverse)
library(chron)
library(here)
library(patchwork)

```

## Read in the data
```{r}

max_temp <- read_csv(here("data", "max_temp_san_joaquin.csv"))
min_temp <- read_csv(here("data", "min_temp_san_joaquin.csv"))
precip <- read_csv(here("data", "precip_san_joaquin.csv"))

#need to tidy to separate out the year and model 

max_tidy <- data.frame(max_temp) %>% 
  rename("tmax" = "value") %>% 
  select(-min, -max) %>% 
  mutate(date = str_split(date, pattern = "00:00:00")) %>% 
  unnest(date) %>% 
  mutate(date = str_trim(date)) %>% 
  filter(!date == "GMT-0800 (Pacific Standard Time)") %>% 
  filter(!name == "Observed") %>% 
  filter(name == "HadGEM2-ES") %>%
  mutate(date = str_sub(date, start = -4)) %>% 
  mutate(date = as.numeric(date)) %>% 
  filter(date %in% c(2020:2099)) %>% 
  mutate(tmax = (tmax-32)*(5/9))

min_tidy <- data.frame(min_temp) %>% 
  rename("tmin" = "value") %>% 
  select(-min, -max) %>% 
  mutate(date = str_split(date, pattern = "00:00:00")) %>% 
  unnest(date) %>% 
  mutate(date = str_trim(date)) %>% 
  filter(!date == "GMT-0800 (Pacific Standard Time)") %>% 
  filter(!name == "Observed") %>% 
  filter(name == "HadGEM2-ES") %>%
  mutate(date = str_sub(date, start = -4)) %>% 
  mutate(date = as.numeric(date)) %>% 
  filter(date %in% c(2020:2099)) %>% 
  mutate(tmin = (tmin-32)*(5/9))

precip_tidy <- data.frame(precip) %>% 
  rename("precip" = "value") %>% 
  select(-min, -max) %>% 
  mutate(date = str_split(date, pattern = "00:00:00")) %>% 
  unnest(date) %>% 
  mutate(date = str_trim(date)) %>% 
  filter(!date == "GMT-0800 (Pacific Standard Time)") %>% 
  filter(!name == "Observed") %>% 
  filter(name == "HadGEM2-ES") %>%
  mutate(date = str_sub(date, start = -4)) %>% 
  mutate(date = as.numeric(date)) %>% 
  filter(date %in% c(2020:2099)) %>% 
  mutate(precip = precip/12)

```

## Explore the data, visually explore the data
```{r}

#plot the data
temp_plot <- ggplot()+
  geom_line(data = max_tidy, aes(x = date, y = tmax),
            color = "firebrick3") +
  geom_line(data = min_tidy, aes(x = date, y = tmin),
            color = "dodgerblue3") +
  theme_minimal() +
  labs(title = "Modeled Annual Max and Min Temps",
       x = "Year",
       y = "Temp (°C)")

precip_plot <- ggplot(precip_tidy, aes(x = date, y = precip))+
  geom_line(color = "slateblue4") +
  theme_minimal() +
  labs(title = "Modeled Monthly Precipitation",
       x = "Year",
       y = "Precipitation (inches)")

temp_precip <- temp_plot+precip_plot 
  
  
temp_precip+
  plot_annotation(
    title = 'Annual Averages for the San Joaquin County, California (2020-2099)',
    theme = theme(plot.title = element_text(size = 15,
                                            face = "bold",
                                            hjust = 0.5)))


```

##Add the function from the empirical model for ORANGES and AVOS
```{r}

# Make a new, combined data frame for ORANGES
combined_orange <- precip_tidy %>% 
  full_join(max_tidy) %>% 
  full_join(min_tidy) %>% 
  mutate(yield = 1.08*tmin -0.20*(tmin^2)+4.99*precip-1.97*(precip^2)-2.47)

# Make a new, combined data frame for AVOCADOS
combined_avos <- precip_tidy %>% 
  full_join(max_tidy) %>% 
  full_join(min_tidy) %>% 
  mutate(yield = 17.71*tmax -0.29*(tmax^2)+3.25*tmin-0.14*(tmin^2)+1.0*precip-0.31*(precip^2)-288.09)

#plot them
ggplot() +
  geom_jitter(data = combined_orange, aes(x = date, y = yield),
              color = "darkorange2") +
  geom_smooth(data = combined_orange, aes(x = date, y = yield),
              color = "darkorange2",
              alpha = .5) +
  geom_jitter(data = combined_avos, aes(x = date, y = yield),
              color = "chartreuse4") +
  geom_smooth(data = combined_avos, aes(x = date, y = yield),
              color = "chartreuse4",
              alpha = .5) +
  theme_minimal() +
  labs(title = "Modeled Yield Anomalies for Oranges and Avocados",
       subtitle = "San Joaquin County, California (2020-2099)",
       x = "Year", 
       y = "Yield (% Anomaly)") +
  theme(plot.title = element_text(hjust = 0.5, margin=margin(0,0,15,0)),
        plot.subtitle = element_text(hjust = 0.5, margin=margin(0,0,15,0)))

```
